### 2019-03-22 paper/sex_selection_spacing-ver2.tex
Grammar edits of Introduction

Birth spacing has long served as a measure of son preference, with strong
son preference typically associated with shorter birth intervals the
fewer sons a family has \citep{ben-porath76b,Leung1988}.
However, the introduction of prenatal sex determination has the potential to
fundamentally change the relationship between son preference and birth spacing.
The stronger the son preference, the more likely couples are to resort to sex-selective 
abortions, and each abortion automatically increases the interval between births by six 
months to a year.%
\footnote{
The increase consists of three parts.
First, starting from the time of the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Longer birth intervals in the absence of sons, which previously would be taken to indicate 
a lower son preference, now may arise precisely because a sufficiently strong son 
preference makes couples use sex selection.

@@

% [Why spacing?] 
There are four main reasons for examining birth spacing and how it changes with 
access to sex selection.
First, researchers have made extensive use of birth spacing as a measure of son preference, 
and it is critical to understand to what extent spacing is still useful as a measure of son 
preference.
In India, the sex of the last-born child significantly impacts the duration to next birth,
with the expected birth interval about three percent longer if the last-born was a son than
if it was a daughter \citep{Bhalotra2008,Kumar2016}.
Similarly, ethnic Indians in South Africa also show a longer duration after the birth of a 
son than after a daughter \citep{Gangadharan2003}.%
\footnote{
Ethnic Indians in Malaysia do, however, show little evidence of son preference, although
the sample sizes are small \citep{Pong1994}.
}
In both Bangladesh and Vietnam the more boys a family has the longer the expected duration 
to the next birth \citep{Haughton1995,Haughton1996,Rahman1993,Soest2018}.
Outside Asia, the evidence is more mixed.
North Africa shows shorter spacing in the absence of sons, while a similar pattern does
not exist in Sub-Saharan Africa \citep{Rossi2015}.%
\footnote{
A possible exception is Senegal, where women with a higher risk of widowhood show
shorter birth spacing until they have secured a son \citep{Lambert2016}.
}

@@

% [Hypotheses / questions]
Despite the increasing use of sex selection, the important impacts of spacing on outcomes 
for both mother and children, and the extensive use of spacing to measure son 
preference in the literature, there has so far been no examination of what effect access 
to sex selection has had on spacing.
The overall hypothesis is that spacing patterns significantly changed as prenatal sex 
determination became available.
Specifically, I address three questions.

First, how did average birth intervals change over time for different groups and how are 
those changes related to the likely use of sex selection?
The underlying hypothesis is that in the absence of sex selection son preference leads to 
shorter spacing the fewer boys a family has for a given parity, whereas son preference 
increases the spacing the fewer boys a family has when prenatal sex selection is available.
I still expect parents to \emph{try} to conceive earlier after the birth of a girl than 
after a boy, but the abortions will increase the duration between births, and may even 
lead to longer average spacing after the birth of girls than after boys.
Families with son preference who---for one reason or another---do not use prenatal sex 
selection may continue to have shorter birth spacing.
Based on the prior literature on sex selection, I focus on how birth intervals have
changed for different education levels, whether urban or rural, and the sex composition of 
prior children.

Second, have the shapes of the survival curves changed?
Very short birth intervals are associated with worse health outcomes and increasing 
mortality risk \citep{Conde-Agudelo2012}.
Hence, understanding how the distribution of birth intervals changes help us understand to 
what extent we can expect improvements in health outcomes.
We should, for example, expect different impacts on the distribution of birth intervals
depending on whether parents change their spacing behavior based on an increasing 
concern for child health or from a strong son preference leading to continued pregnancies
until a son is conceive.

@@

Direct information on the use of sex selection is not available, but it is possible
to compare different periods as a proxy since there has been substantial changes in access 
and legality of prenatal sex determination in India.
Abortion has been legal in India since 1971 and still is.
The first reports of sex determination appeared around 1982--83 
\citep{Sudha1999,bhat06,Grover2006}.
The number of clinics quickly increased, and knowledge about sex selection became 
widespread after a senior government official's wife aborted a fetus that turned out to be 
male \citep[p.\ 598]{Sudha1999}.
In 1994, the Central Government passed the Prenatal Diagnostic Techniques (PNDT) Act, 
making determining and communicating the sex of a fetus illegal.%
\footnote{
Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
The number of convictions has been low.
It took until January 2008 for the first state, Haryana, to reach five convictions.
Hence, private clinics apparently operate with little risk of legal action 
\citep{Sudha1999}.
Furthermore, there is little evidence that bans like this significantly
affect sex ratios \citep{Das-Gupta2016}.
}

Based on these development, I split the four decades into four periods.
The first period cover until prenatal sex determination became widely available, 
1972--1984.
The second period saw (mostly) unrestricted access to sex selection, 1985--1994.
Although, as reviewed above, there is evidence that there has been a substantial 
increase in the use of sex selection, some researchers argue that we might be close to or 
have already passed a turning-point in the use of sex selection 
\citep{Das_Gupta2009,Diamond-Smith2015}.
I, therefore, split the period from the implementation of the PNDT Act until the last 
survey into two, 1995--2004, and 2005--2016.

There are four main results.
First, there has been a general increase in the intervals between births for all education 
groups over the four decades covered by the data.
These increases are larger the higher the parity and the higher the education level.
Second, those women who are most likely to use sex selection, well-educated women with no
sons, have seen the largest increases in birth intervals.
As a result, we now have a complete reversal of the traditional spacing pattern for
well-educated women, and instead observe the longest spacing for women with no sons.
Third, women who are least likely to use sex selection, those with no education in
rural areas, still follow the standard pattern of short spacing when they have girls and 
little evidence of the use of sex selection.
In other words, these women adhere to a strategy where they achieve a son through higher 
fertility rather than the use of sex selection.
Fourth, the traditional pattern of longer average spacing when women have one or more sons
appear to arise from more very long intervals, rather than from fewer very short birth 
intervals; across education groups there appear to be little difference in the likelihood
of having a very short birth interval across sex compositions, especially for lower
parities.

>>>>

Birth spacing has long served as a measure of son preference, with strong
son preference typically associated with shorter birth intervals the
fewer sons a family has \citep{ben-porath76b,Leung1988}.
However, the introduction of prenatal sex determination has the potential to
change the relationship between son preference and birth spacing.
The stronger the son preference, the more likely couples are to resort to sex-selective 
abortions, and each abortion automatically increases the interval between births by six 
months to a year.%
\footnote{
The increase consists of three parts.
First, starting from the time of the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Longer birth intervals in the absence of sons, which previously would be taken to indicate 
a lower son preference, now may arise precisely because a sufficiently strong son 
preference makes couples use sex selection.

@@

% [Why spacing?] 
There are four main reasons for examining birth spacing and how it changes with 
access to sex selection.
First, researchers have made extensive use of birth spacing as a measure of son preference, 
and it is critical to understand to what extent spacing is still useful as a measure of son 
preference.
In India, the sex of the last-born child significantly impacts the duration to next birth,
with the expected birth interval about three percent longer if the last-born was a son than
if it was a daughter \citep{Bhalotra2008,Kumar2016}.
Similarly, ethnic Indians in South Africa also show a longer duration after the birth of a 
son than after a daughter \citep{Gangadharan2003}.%
\footnote{
Ethnic Indians in Malaysia do, however, show little evidence of son preference, although
the sample sizes are small \citep{Pong1994}.
}
In both Bangladesh and Vietnam the more boys a family has the longer is the expected 
duration to the next birth \citep{Haughton1995,Haughton1996,Rahman1993,Soest2018}.
Outside Asia, the evidence is more mixed.
North Africa shows shorter spacing in the absence of sons, while a similar pattern does
not exist in Sub-Saharan Africa \citep{Rossi2015}.%
\footnote{
A possible exception is Senegal, where women with a higher risk of widowhood show
shorter birth spacing until they have secured a son \citep{Lambert2016}.
}

@@

% [Hypotheses / questions]
Despite the increasing use of sex selection, the critical impacts of spacing on outcomes 
for both mother and children, and the extensive use of spacing to measure son 
preference in the literature, there has so far been no examination of what effect access 
to sex selection has had on spacing.
The overall hypothesis is that spacing patterns significantly changed as prenatal sex 
determination became available.
Specifically, I address three questions.

First, how did average birth intervals change over time for different groups and how are 
those changes related to the likely use of sex selection?
The underlying hypothesis is that in the absence of sex selection son preference leads to 
shorter spacing the fewer boys a family has for a given parity, whereas son preference 
increases the spacing the fewer boys a family has when prenatal sex selection is available.
I still expect parents to \emph{try} to conceive earlier after the birth of a girl than 
after a boy, but the abortions will increase the duration between births, and may even 
lead to longer average spacing after the birth of girls than after boys.
Families with son preference who---for one reason or another---do not use prenatal sex 
selection may continue to have shorter birth spacing.
Based on the prior literature on sex selection, I focus on how birth intervals have
changed for different education levels, urban or rural, and the sex composition of 
prior children.

Second, have the shapes of the survival curves changed?
Very short birth intervals are associated with worse health outcomes and increased 
mortality risk \citep{Conde-Agudelo2012}.
Hence, understanding how the distribution of birth intervals changes help us understand to 
what extent we can expect improvements in health outcomes.
We should, for example, expect different impacts on the distribution of birth intervals
depending on whether parents change their spacing behavior based on an increasing 
concern for child health or from a strong son preference leading to continued pregnancies
until they conceive a son.

@@

Direct information on the use of sex selection is not available, but it is possible to 
compare different periods as a proxy since there have been substantial changes in access 
and legality of prenatal sex determination in India.
Abortion has been legal in India since 1971 and still is.
The first reports of sex determination appeared around 1982--83 
\citep{Sudha1999,bhat06,Grover2006}.
The number of clinics quickly increased, and knowledge about sex selection became 
widespread after a senior government official's wife aborted a fetus that turned out to be 
male \citep[p.\ 598]{Sudha1999}.
In 1994, the Central Government passed the Prenatal Diagnostic Techniques (PNDT) Act, 
making determining and communicating the sex of a fetus illegal.%
\footnote{
Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
The number of convictions has been low.
It took until January 2008 for the first state, Haryana, to reach five convictions.
Hence, private clinics operate with little risk of legal action \citep{Sudha1999}.
Furthermore, there is little evidence that bans like this significantly
affect sex ratios \citep{Das-Gupta2016}.
}

Based on these developments, I split the four decades into four periods.
The first period covers until prenatal sex determination became widely available, 
1972--1984.
The second period saw (mostly) unrestricted access to sex selection, 1985--1994.
Although, as reviewed above, there is evidence that there has been a substantial 
increase in the use of sex selection, some researchers argue that we might be close to or 
have already passed a turning-point in the use of sex selection 
\citep{Das_Gupta2009,Diamond-Smith2015}.
I, therefore, split the period from the implementation of the PNDT Act until the last 
survey into two, 1995--2004, and 2005--2016.

There are four main results.
First, there has been a general increase in the intervals between births for all education 
groups over the four decades covered by the data.
These increases are larger the higher the parity and the higher the education 
level.
Second, those women who are most likely to use sex selection, well-educated women with no
sons, have seen the largest increases in birth intervals.
As a result, we now have a complete reversal of the traditional spacing pattern for
well-educated women, and instead, observe the longest birth intervals for women with no 
sons.
Third, women who are least likely to use sex selection, those with no education in
rural areas, still follow the standard pattern of short spacing when they have girls and 
little evidence of the use of sex selection.
In other words, these women adhere to a strategy where they achieve a son through higher 
fertility rather than the use of sex selection.
Fourth, the traditional pattern of longer average spacing when women have one or more sons
appear to arise from more very long intervals, rather than from fewer very short birth 
intervals; across education groups there appear to be little difference in the likelihood
of having a very short birth interval across sex compositions, especially for lower
parities.
